An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-18 thesis/compact-finite-difference.tex
Corrections

Here, we consider the case of multiple GPUs
on a single shared memory node, i.e.,
multiple GPUs attached to the same PCI-e bus.
In this case, every GPU is visible to the host,
and the GPUs read from and write into
the same host memory space.
The domain is divided into a number of
into ``subdomains'',
as shown in \ref{fig:compact-shared-gpu}.
The domain decomposition is done such that
only a single subdomain is used along the
coordinate direction of the derivatives.
This is the method presented by
Sakharnykh et al. \cite{sakharnykhADIconf},
and it has the advantage that
no coordination between GPUs is required:
each GPU is assigned an independent
set of grid lines to solve.

@@

For larger problems,
a distributed system may be required.
Here, the simplest strategy is to use a domain decomposition
as shown in \ref{fig:compact-distributed-restricted}.
Here again, 
no inter-GPU communication is required.
By restricting the distribution along one coordinate direction,
the ease of solution of the tridiagonal systems is maintained.

@@

\item For domains that are much longer along one coordinate
    direction compared to the other(s),
    the subdomains may become impractically \emph{slender}.
\end{enumerate}
%
Such decomposition strategies are therefore, generally applicable
when compact finite difference schemes are used only
in a single direction.
For the other coordinate directions,
explicit finite difference schemes may be used,
which have the disadvantages discussed earlier.
To accommodate compact finite difference schemes in all the coordinate
directions,
the domain decomposition generally must be performed in all directions,
as shown in Fig. \ref{fig:compact-distributed-all}.
In this strategy,
the grid lines in all coordinate directions
are interrupted by the subdomain boundaries.
Thus,
inter-GPU communication is required.
For the right hand sides evaluation,
each GPU must communicate information at the subdomain boundaries
with neighbouring GPUs.
For example, in Fig. \ref{fig:compact-distributed-all},
the right-most grid points in the subdomain of
GPU 2 require data from the
left-most grid points in the subdomain of
GPU 6 when evaluating derivatives in the horizontal direction.
Similarly,
the left-most grid points in the subdomain of
GPU 6 require data from the
right-most grid points in the subdomain of
GPU 2.
Thus, a ``swapping'' of the boundary information
is required at each of the subdomain boundaries.

The solution of the distributed tridiagonal systems
involves much more complexity,
and is discussed in detail in the next section.

@@

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
\end{align}
%
\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2-expanded}
\end{align}
%
We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
the ``primary'' system, and the subsystems in
Eqs. (\ref{eqn:secondary-system-1-expanded}) and
(\ref{eqn:secondary-system-2-expanded})
the ``secondary'' systems.
Additionally, the following ``reduced'' system is solved:

@@

Our algorithm is outlined in \ref{fig:compact-algorithm},
and is described for the case of evaluating derivatives
in the direction along which successive
function values in a subdomain are stored contiguously in memory,
i.e., the ``fastest'' coordinate direction.
Each step of the algorithm must be implemented on the GPU
to avoid data transfer to and from the CPU,
which is prohibitively expensive for large problems.
Thus, we have several kernels to implement the algorithm.
For communication of data between processes,
we use the Message Passing Interface (MPI),
and leverage the NVIDIA GPUDirect Technology
for GPU-GPU communication.
MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
Python library.
The purpose of each CUDA kernel and MPI call used
is described in Table \ref{table:compact-algorithm}.

@@

\begin{table}[]
\centering
\caption{Purpose of kernels and MPI calls in compact finite difference application}
\label{table:compact-algorithm}
\begin{tabular}{|l|p{6cm}|}
\hline
Memcpy3d            & Copy noncontiguous boundary information
                      of the function values
                      to and from contiguous halo arrays \\ \hline
ISend, IRecv        & Perform halo swaps with i-1 and i+1 processes \\ \hline
computeRHSKernel    & Apply pointwise stencil operator to compute
                      RHS at each grid point in the subdomain,
                      using the halo values near the boundaries\\ \hline
NEATO               & Solve the primary near-Toeplitz tridiagonal systems
                      for the computed right hand sides, giving $x_r$ \\ \hline
copyFacesKernel     & Copy the left and right faces of $x_r$
                      into a single contiguous array \\ \hline
MPI\_Gather         & Gather the data required to assemble
                      the reduced system at rank 0 \\ \hline
reducedSolverKernel & Solve the reduced systems for parameters
                      $\alpha^p$ and $\beta^p$
                      for each grid line \\ \hline
MPI\_Scatter        & Scatter the parameters 
                      $\alpha^p$ and $\beta^p$
                      from rank 0 to all the processes \\ \hline
sumSolutionsKernel  & Sum the primary and secondary solutions
                      to compute the local part of the solution \\ \hline
\end{tabular}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=200pt]{img/constructing-reduced-system.eps}
\caption{Construction of the reduced system and scattering of parameters.}
\label{fig:constructing-reduced-system}
\end{center}
\end{figure}
%
The secondary systems
[Eq. (\ref{eqn:secondary-system-1-expanded}) and
 Eq. (\ref{eqn:secondary-system-2-expanded})]
are solved on the CPU,
and the results are transferred to the GPU.
%
The primary system [Eq. (\ref{eqn:primary-system-expanded})] must be solved
for each of the local grid lines in a subdomain,
as the right hand sides are different at each of the local grid lines. 
The evaluation of the right hand sides are
pointwise stencil computations,
which require communication of the
function values at the boundaries of the subdomains.
This is achieved using a halo-swapping technique
with dedicated contiguous halo arrays for each subdomain,
as communication of non-contiguous MPI
data types is expensive on the GPU.
The primary system is solved for all the local grid lines
using the NEATO solver.
%
The set up and solution of the reduced system
requires global communication of the boundary information
from the solutions $\bm{u}^p$, $\bm{l}^p$ and $\bm{x_r}^p$.
The tridiagonal coefficients of the reduced system
are set up easily,
by communicating the boundary elements from
$\bm{u}^p$ and $\bm{l}^p$,
which are the same for every local grid line in each subdomain.
The right hand sides require significantly more communication,
as they are assembled from the boundary elements of $\bm{x_r}^p$,
which are different for each local grid line in each subdomain.
Thus, the boundary ``faces'' of each subdomain need to be communicated.
These faces are first copied into a contiguous array,
and these arrays are gathered at rank 0
to assemble the right-hand sides of the reduced system
(Fig. \ref{fig:constructing-reduced-system}).
This communication strategy has the effect of
producing \emph{interleaved} right hand sides
aligned along the ``slowest'' co-ordinate direction, i.e.,
the right hand side values for neighbouring grid points
are located far apart in memory.
As the reduced systems are quite small relative to the primary systems,,
we use the convenient, but rather inefficient
p-Thomas algorithm to solve the systems.
The solution of the reduced systems produces the parameters
$\alpha^p$ and $\beta^p$,
which are scattered back to the respective processes, $p$.
%
Finally, the summing of the solutions
is a pointwise operation that is easily implemented
on the GPU.
For evaluation of compact finite differences in other
coordinate directions,
a local permutation of the data is performed on the input data
(function values)
before applying the above algorithm.
and again on the output data
(derivative values).

>>>>

Here, we consider the case of multiple GPUs
on a single shared memory node, i.e.,
multiple GPUs attached to the same PCI-e bus.
In this case, every GPU is visible to the host,
and the GPUs read from and write into
the same host memory space.
The domain is divided into a number of `subdomains'',
as shown in \ref{fig:compact-shared-gpu}.
The domain decomposition is done such that
only a single subdomain is used along the
coordinate direction of the derivatives.
This is the method presented by
Sakharnykh et al. \cite{sakharnykhADIconf},
and it has the advantage that
no coordination between GPUs is required:
each GPU is assigned an independent
set of grid lines to solve.

@@

For larger problems,
a distributed system is nearly always required.
Here, the simplest strategy is to use a domain decomposition
as shown in Fig. \ref{fig:compact-distributed-restricted}.
Here again, 
no inter-GPU communication is required.
By restricting the distribution along one coordinate direction,
the ease of solution of the tridiagonal systems is maintained.

@@

\item For domains that are much longer along one coordinate
    direction compared to the other(s),
    the subdomains may become impractically \emph{slender}.
\end{enumerate}
%
Such decomposition strategies are therefore, generally applicable
when compact finite difference schemes are used only
in a single direction.
For the other coordinate directions,
explicit finite difference schemes may be used,
which have the disadvantages discussed earlier.
To accommodate compact finite difference schemes in all the coordinate
directions,
the domain decomposition generally must be performed in all directions,
as shown in Fig. \ref{fig:compact-distributed-all}.
In this strategy,
the grid lines in all coordinate directions
are interrupted by the subdomain boundaries.
Thus,
inter-GPU communication is required.
For the right hand sides evaluation,
each GPU must communicate information at the subdomain boundaries
with neighbouring GPUs.
For example, in Fig. \ref{fig:compact-distributed-all},
the right-most grid points in the subdomain of
GPU 2 require data from the
left-most grid points in the subdomain of
GPU 6 when evaluating derivatives in the horizontal direction.
Similarly,
the left-most grid points in the subdomain of
GPU 6 require data from the
right-most grid points in the subdomain of
GPU 2.
Thus, a ``swapping'' of the boundary information
is required at each of the subdomain boundaries.
The solution of the distributed tridiagonal systems
involves much more complexity,
and is discussed in detail in the next section.

@@

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1-expanded} 
\end{align}
%
\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2-expanded}
\end{align}
%
We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
as the ``primary'' system,
and the subsystems in
Eqs. (\ref{eqn:secondary-system-1-expanded}) and
(\ref{eqn:secondary-system-2-expanded})
as the ``secondary'' systems.
Additionally, the following ``reduced'' system
must be constructed andsolved:

@@

Our algorithm is outlined in Fig. \ref{fig:compact-algorithm},
and is described for the case of evaluating derivatives
in the direction along which successive
function values in a subdomain are stored contiguously in memory,
i.e., the ``fastest'' coordinate direction.
Each step of the algorithm must be implemented on the GPU
to avoid data transfer to and from the CPU,
which is prohibitively expensive for large problems.
Thus, we have several kernels to implement the algorithm.
For communication of data between processes,
we use the Message Passing Interface (MPI),
and leverage the NVIDIA GPUDirect Technology
for GPU-GPU communication.
MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
Python library.
The purpose of each CUDA kernel and MPI call used
is described in Table \ref{table:compact-algorithm}.

@@

\begin{table}[]
\centering
\caption{Purpose of kernels and MPI calls in compact finite difference application}

\label{table:compact-algorithm}
\begin{tabular}{|l|p{6cm}|}
\hline
Memcpy3d            & Copy noncontiguous boundary information
                      of the function values
                      to and from contiguous halo arrays \\ \hline
ISend, IRecv        & Perform halo swaps with i-1 and i+1 processes \\ \hline
computeRHSKernel    & Apply pointwise stencil operator to compute
                      RHS at each grid point in the subdomain,
                      using the halo values near the boundaries\\ \hline
NEATO               & Solve the primary near-Toeplitz tridiagonal systems
                      for the computed right hand sides, giving $x_r$ \\ \hline
copyFacesKernel     & Copy the left and right faces of $x_r$
                      into a single contiguous array \\ \hline
MPI\_Gather         & Gather the data required to assemble
                      the reduced system at rank 0 \\ \hline
reducedSolverKernel & Solve the reduced systems for parameters
                      $\alpha^p$ and $\beta^p$
                      for each grid line \\ \hline
MPI\_Scatter        & Scatter the parameters 
                      $\alpha^p$ and $\beta^p$
                      from rank 0 to all the processes \\ \hline
sumSolutionsKernel  & Sum the primary and secondary solutions
                      to compute the local part of the solution \\ \hline
\end{tabular}
\end{table}

\begin{figure}[h]
\begin{center}
\includegraphics[width=200pt]{img/constructing-reduced-system.eps}
\caption{Construction of the reduced system and scattering of parameters.}
\label{fig:constructing-reduced-system}
\end{center}
\end{figure}
%
The secondary systems
[Eq. (\ref{eqn:secondary-system-1-expanded}) and
 Eq. (\ref{eqn:secondary-system-2-expanded})]
are solved on the CPU,
and the results are transferred to the GPU.
%
The primary system [Eq. (\ref{eqn:primary-system-expanded})] must be solved
for each of the local grid lines in a subdomain,
as the right hand sides are different at each of the local grid lines. 
The evaluation of the right hand sides are
pointwise stencil computations,
which require communication of the
function values at the boundaries of the subdomains.
This is achieved using a halo-swapping technique
with dedicated contiguous halo arrays for each subdomain,
as communication of non-contiguous MPI
data types is expensive on the GPU.
The primary system is solved for all the local grid lines
using the NEATO solver.
%
The set up and solution of the reduced system
requires global communication of the boundary information
from the solutions $\bm{u}^p$, $\bm{l}^p$ and $\bm{x_r}^p$.
The tridiagonal coefficients of the reduced system
are set up easily,
by communicating the boundary elements from
$\bm{u}^p$ and $\bm{l}^p$,
which are the same for every local grid line in each subdomain.
The right hand sides require significantly more communication,
as they are assembled from the boundary elements of $\bm{x_r}^p$,
which are different for each local grid line in each subdomain.
Thus, the boundary ``faces'' of each subdomain need to be communicated.
These faces are first copied into a contiguous array,
and these arrays are gathered at rank 0
to assemble the right-hand sides of the reduced system
(Fig. \ref{fig:constructing-reduced-system}).
This communication strategy has the effect of
producing \emph{interleaved} right hand sides
aligned along the ``slowest'' co-ordinate direction, i.e.,
the right hand side values for neighbouring grid points
are located far apart in memory.
As the reduced systems are quite small relative to the primary systems,,
we use the convenient, but rather inefficient
p-Thomas algorithm to solve the systems.
The solution of the reduced systems produces the parameters
$\alpha^p$ and $\beta^p$,
which are scattered back to the respective processes, $p$.
%
Finally, the summing of the solutions
is a pointwise operation that is easily implemented
on the GPU.
For evaluation of compact finite differences in other
coordinate directions,
a local permutation of the data is performed on the input data
(function values)
before applying the above algorithm.
and again on the output data
(derivative values).
